---
title: "Preprocessing"
output: rmarkdown::html_vignette
bibliography: "../bibliography.bib"
vignette: >
  %\VignetteIndexEntry{Preprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

This vignette shows the preprocessing workflow used to analyze raw IDAT files from Illumina HumanMethylationEPIC microarrays.
After reading in the intensity levels, the following steps will be performed:

- calculate detection p-values based on control probes using `minfi::detectionP()`,
- conversion of intensity levels into methylation beta-values and M-values using `minfi::preprocessRaw()`,
- normalization in order to account for the two different probe types employed on the array and other systematic biases of the platform via: 
  - `minfi::preprocessSWAN()`,
  - `minfi::preprocessNoob()`,
  - `minfi::preprocessQuantile()`.
  
The resulting data will be saved to the specified output directory.

After this first preprocessing step, methylation data can be obtained using function `get_beta()` which allows to specify the normalization method, as well as a probe filter and a sample filter.
Probe filtering has been suggested repeatedly in literature since some array probes were found to yield non-reproducible results due to targeting SNPs instead of methylation, non-unique binding sequences or off-target hybridization [@zhou_comprehensive_2016; @pidsley_critical_2016].


```{r setup}
library(intratumormeth)
```

## Data format requirements

To run the preprocessing workflow, a `samplesheet` dataframe has to be prepared which associates a given sample id with the location of corresponding IDAT files.
There should be one green and one red channel IDAT file for each sample which can be distinguished by their file name suffix (e.g. `filename_Red.idat` and `filename_Grn.idat`).
Column `idat_location` should contain the filename _without_ the red/green suffix.
Since this is an _intratumor_ heterogeneity study, the individual samples furthermore have to be associated with corresponding patients.
The patient id corresponding to a given sample should be specified in column `patient_id`.
Additionally, arbitrary sample-wise metadata can be specified by other columns in the samplesheet (take care not to name them ambiguously).

An exemplary samplesheet is shown here:

```{r exemplary samplesheet, echo=FALSE, message=FALSE}
# read idat samplesheet and print head
samplesheet <- read.csv("~/sfb824/packagepdgfra_input/samplesheet.csv")
samplesheet %>% head()
```



This samplesheet contains 213 samples from 46 patients of the SFB824 project B12, as well as 47 samples from 15 patients from the dataset published by @wenger_intratumor_2019 (GEO accession: GSE116298)

## Workflow

Now all required data has been prepared and the actual preprocessing can begin.
To be able to render this vignette faster, only a 10 sample subset of the data will be processed.

```{r, eval=FALSE}
preprocess_methylation_data(samplesheet, outputDir = "~/sfb824/packagepdgfra_output/")
```


Once finished, the generated data can be obtained like so:

```{r, eval=FALSE}
betas <- get_betas(outputDir = "~/sfb824/packagepdgfra_output/", 
                   normalization = "swan", 
                   removeNaProbes = TRUE,
                   probeFilter = NULL,
                   sampleFilter = NULL)
betas[1:5,1:5]
```


Before starting further analysis of these methylation levels, a quality control step should be performed in order to detect samples of poor quality (i.e. high average detection p-value) and filter out probes that are associated with low reproducibility (TODO link to Zhou et al. and Pidsley et al.).
See vignette `quality_control` for further details.























## Bibliography




